---
title: "STOR 565 Final Project"
author: "Group 6: Vikram P.T., Suyang Wang, Rui Li, Tyler Youngberg"
date: "4/19/2020"
output: pdf_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Refered to https://datascienceplus.com/fitting-neural-network-in-r/ and https://datascienceplus.com/neuralnet-train-and-test-neural-networks-using-r/

```{r}
library(MASS)
library(neuralnet)
library(corrplot)
library(MLmetrics)
```

```{r}
wine <- read.csv('Combined Wine.csv')

##wine type: Red - 1; White -2##
wine$Type <- as.factor(as.numeric(wine$Type))


wine <- wine[,-c(1,13)]
head(wine)
set.seed(233)

## Checking for missing data##
apply(wine, 2, function(x) sum(is.na(x)))  # no missing data
```


```{r}
sum = summary(wine)

ggplot(wine,aes(Type,total.sulfur.dioxide,fill=Type))+geom_boxplot(outlier.colour = "red") + labs(x="Wine Type", y="Total Sulphur Dioxide")+ggtitle("Total Sulphur Dioxide")+theme(plot.title = element_text(size = 30,hjust = 0.5, face = "bold"))

ggplot(wine,aes(Type,alcohol,fill=Type))+geom_boxplot(outlier.colour = "red") + labs(x="Wine Type", y="Alcohol")+ggtitle("Alcohol")+theme(plot.title = element_text(size = 30,hjust = 0.5, face = "bold"))

ggplot(wine,aes(Type,pH,fill=Type))+geom_boxplot(outlier.colour = "red") + labs(x="Wine Type", y="pH")+ggtitle("PH")+theme(plot.title = element_text(size = 30,hjust = 0.5, face = "bold"))

ggplot(wine,aes(Type,residual.sugar,fill=Type))+geom_boxplot(outlier.colour = "red",outlier.shape = NA) + labs(x="Wine Type", y="residual.suagr")+scale_y_continuous(breaks = seq(0, 30, 5),limits=c(0, 30))+ggtitle("Residual Sugar")+theme(plot.title = element_text(size = 30,hjust = 0.5, face = "bold"))
```



```{r}
corrplot(cor(wine[1:11]),type="lower")  # Alcohol + Sugar + Water = Density Thats density and sugar and density and alcohol might be highly correlated
# Total SO2= BoundSO2 + Free SO2

pairs(wine[1:11])
```


Fact: Red wines typically contain less SO2 than white wine.

Why do red wines have less sulfites (S02)? They contain tannin, which is a stabilizing agent, and almost all red wines go through malolactic fermentation. Therefore, less sulfur dioxide is needed to protect the wine during winemaking and maturation.

```{r}
ggplot(wine)+geom_density(aes(x=total.sulfur.dioxide,color=Type))

mean(wine$total.sulfur.dioxide~wine$Type)
```





```{r}
combined_pcs<-prcomp(wine[,1:11])

summary(combined_pcs)

screeplot(combined_pcs,type = "lines")

```


Trying to capture max variance in lower dimension. PC1 plotted again 4,6,7 shows separability in the type of wine.

```{r}
ggplot(as.data.frame(combined_pcs$x))+geom_point(aes(x=combined_pcs$x[,4],y=combined_pcs$x[,1],color=wine$Type))+labs(x="PC4",y="PC1",title = "Plot of data projected onto PCs",color="Wine Type")+theme(plot.title = element_text(size = 30,hjust = 0.5, face = "bold"))

ggplot(as.data.frame(combined_pcs$x))+geom_point(aes(x=combined_pcs$x[,6],y=combined_pcs$x[,1],color=wine$Type))+labs(x="PC6",y="PC1",title = "Plot of data projected onto PCs",color="Wine Type")+theme(plot.title = element_text(size = 30,hjust = 0.5, face = "bold"))

ggplot(as.data.frame(combined_pcs$x))+geom_point(aes(x=combined_pcs$x[,7],y=combined_pcs$x[,1],color=wine$Type))+labs(x="PC7",y="PC1",title = "Plot of data projected onto PCs",color="Wine Type")+theme(plot.title = element_text(size = 30,hjust = 0.5, face = "bold"))

combined_proj<-combined_pcs$x[,1:7]
```




CREATING THE MODEL



```{r}
# Splitting data into training and testing set
index <- sample(1:nrow(wine),round(0.75*nrow(wine)))
maxs <- apply(wine[1:11], 2, max) 
mins <- apply(wine[1:11], 2, min)
scaled <- as.data.frame(scale(wine[1:11], center = mins, scale = maxs - mins))
training_set <- scaled[index,]
training_set$Type<-ifelse(wine$Type[index]==1,TRUE,FALSE)
testing_set <- scaled[-index,]
testing_set$Type<-ifelse(wine$Type[-index]==1,TRUE,FALSE)

head(training_set)
```


```{r}
# fit the net

nn <- neuralnet(Type~., data=training_set, hidden=c(5,3), linear.output=F, err.fct = "ce",likelihood = TRUE, act.fct = "logistic")
```



```{r}
plot(nn)
```




```{r}
nn.results <- compute(nn,testing_set[,1:11])
results <- data.frame(actual = testing_set$Type, prediction = nn.results$net.result)
roundedresults <- sapply(results, round, digits=0)
roundedresultsdf=data.frame(roundedresults)
attach(roundedresultsdf)
table(actual,prediction)
```

Accuracy Rate: 1611/1624 = 99.2%

```{r}
pred= predict(nn,testing_set[,1:11])
class_pred= (pred>=0.5)*1  
res<- data.frame(real=testing_set$Type*1,predicted=class_pred)
head(res)
table((testing_set$Type)*1,class_pred)

```

```{r}
for (i in 1:nrow(sta)) {
 sum = summary(wine[,i])
 sta$Min.[i]=sum[1]
 sta$`1st Qu`[i]=sum[2]
 sta$Median[i]=sum[3]
 sta$Mean[i]=sum[4]
 sta$`3rd Qu`[i]=sum[5]
 sta$Max.[i]=sum[6]
}
```

